---
title: "Quantifying Stability and Change in Personal Culture Using Panel Data"
author:
- 
output:
  pdf_document:
    keep_tex: true
bibliography: kileybib.bib
csl: asa.csl
abstract: "Recent work has produced conflicting interpretations on the question of whether adults undergo durable changes in their personal culture over time. This paper asserts that this divergence is less due to their findings than the approach taken -- a tournament of models focusing on the presence or absence of change. To move beyond this, we reanalyze the data used previously in this debate with an approach that quantifies the processes of cultural difference. In doing so, we compare how much cultural difference is explained by intrapersonal change and interpersonal differences. Our results harmonize recent findings by showing that, while most measures of personal culture show change over time, the variance explained by intrapersonal change is typically small compared to interpersonal differences at baseline. Quantifying these processes offers new ways to address theoretical questions of when and how their relative importance shifts. We examplify this with the case of college experience."
fontsize: 11pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, message = F, warning = F, fig.align = "center")
```

# Introduction

A recurring focus in the sociology of culture is the question of whether people, as they move through their lives and transition between social roles, change their personal culture –- the declarative and non-declarative attitudes, beliefs, values, practices, and dispositions [@lizardo2017; @kiley2020; @lersch2023]. 

This seemingly simple question underlies a number of central debates. For example, pragmatist theories of action claim that changes in social environments cause people to adapt their views and make new meanings [@swidler2001; @gross2009], while Bourdieusian practice theories argue that the “past conditions of production” leave a mark on people's personal culture that lasts throughout their lives [@bourdieu1990]. Similarly, models of social influence assume that people adapt their culture in the face of new information [@vaisey2010a], while the emphasis on cohort effects in models of aggregate social and cultural change requires them to be open to change while young but become fairly resistant to it as they age [@ryder1965]. Finally, life course theories posit slower but still important changes over time as people advance through important transitions in their lives [@elder2016; @bardi2011].

All of the above theoretical perspectives have some empirical support. All the more it is important to adjudicate when and how they apply to advance an understanding the differences in people's personal culture that we observe. That being said, scholars have struggled to reach a consensus on the relative importance of change during adulthood for explaining cultural differences. Over the span of a few years, most apparent changes in people’s responses to personal culture items appear to be transitory, with little evidence of persistent change among adults [@kiley2020; @vaisey2021]. When looked at over longer time horizons, however, there is at least some evidence that adults make persistent cultural changes, especially as they move through distinct social roles across the life course [@lersch2023].

These seemingly inconsistent findings are partly due to the fact that researchers have taken a “tournament of models” approach [@lersch2023: p. 228] in adjudicating the different perspectives. Accordingly, researchers have formalized models of responses to cultural items over time as demonstrating either “change” or “no change” to then compare which model fits observed data better, thus functionally asking whether we ever observe within-person change in personal culture.

We argue, however, that asking only *whether* people change obscures what are actually highly compatible results. Whether change “really happened” in a given data set depends not only on sample populations, item measurement, statistical methods and power, but crucially also on definitions of what counts as “change,” thus running the risk of getting caught up in unresolvable debates about related decisions. [MAYBE BETTER TO SAY HERE: WHICH QUICKLY RELEGATES THE POSSIBILITY OF AN ANSWER BACK TO AN ULTIMATELY UNRESOLVABLE THEORETICAL DEBATE.] Moreover, it narrows the focus on particular questions such as, “Do we see change in this measure?” Such inquiries are ill-suited to address what’s at the core of the theoretical debate. There is enough evidence to assume that neither socialization at an early age nor the influence of the contemporary environment alone can explain cultural differences. The question that remains is under what conditions, across which groups, and in what domains do we observe differences in the relative importance of intrapersonal change and interpersonal differences. This focus is necessarily comparative and requires an approach that quantifies forms of cultural change rather than calling victory for the one or the other perspective.

In this paper, we try to move beyond previous debates by moving away from the “tournament of models” approach toward a quantification of the processes that lead to differences in personal culture. Drawing on the seven panel data sets analyzed in these debates, we use established methods to quantify the amount of variance in responses that is due to between-people differences in baselines versus the amount of variance in responses that is due to linear within-person change over time. This measure gives us a sense of the importance of intrapersonal change over the course of the panel relative to the differences that exist between people when they start. We then explore features of questions and populations that explain differences in within-person change over time.

We observe similar patterns across the data sets that have produced conflicting interpretations. In all data sets we see similar rates of individual-level change over time across a broad range of measures of personal culture. But quantifying these processes also reveals that intrapersonal change plays a substantially smaller role in explaining the variance in peoples’ personal culture than between-person differences in their baselines. Of course, whether this amount of intrapersonal change is “meaningful” ultimately remains a theoretical question reflected in prior expectations concerning the question, population, and time. Nevertheless, the results suggest that the intrapersonal change observed (in some cases over more than 10 years) does not seem to play a substantial role in explaining personal cultural differences between adults. 

This pattern of results helps clarify and, hopefully, transcend previous debates. What is more, the measure we propose offers a quantity with more diverse applications than a simple judgment about whether change happened. To examplify this, we apply the measure to explore differences in the relative importance of intrapersonal change for people with and without college degrees, comparing both within and across questions in ways that the tournament of models cannot. 

While adding to existing debates, we hope the quantification we propose extends them in new ways and enables scholars to more clearly adjudicate theoretical mechanisms at play in shaping people's personal culture.


# Background

## Change and Stability in Personal Culture

Recent debates about whether adults undergo intrapersonal cultural change emerged in part because sociological models of cultural change at the aggregate level tend to implicitly invoke one of two models of individual behavior. The first, what Kiley and Vaisey [-@kiley2020] call a “Settled Dispositions Model” (SDM), assumes that peoples’ personal culture is relatively fixed by the time they are adults. While they might make temporary changes in their declarative culture as their social environments provide new considerations, this model assumes that they return to a settled baseline over a (relatively) short time frame. This model of behavior is frequently invoked in models of cultural change that suggest people are strongly imprinted by the “past conditions of production” of socialization, such as Bourdieusian practice theories [@bourdieu1990], cohort replacement theories of aggregate change [@mannheim1952; @ryder1965], and control theories in social psychology [@smith-lovin1988; @robinson2007]. 

The second model summarized by Kiley and Vaisey [-@kiley2020], an “Active Updating Model” (AUM), posits that people continually update their personal culture as they move through life. This model suggests people make new cultural meanings, and therefore durably change their personal culture, as they encounter new social environments, discourses, and information [@gross2009; @swidler2001]. This model underlies models of cultural diffusion [@christakis2010], attitude alignment [@dellaposta2020], and polarization [@bail2018], among others. It is also implicit in most studies of whether specific experiences, changes in social roles, or political events, affect personal culture [@gelman2021; @visser2004; @slothuus2021].

There is no reason to think that only one of these models is “correct” in all times and places. A population observed over some time frame contains a mix of people who are changing and people who are not (or some people making very small or imperceptible changes in culture and others making perceptible changes). Instead, different perspectives argue that each of these ideal-typical models is more operative at different times, for different people, and for different elements of personal culture. For example, adolescence and early adulthood, typically viewed as a “formative period” for personal cultural development, is likely characterized by higher rates of active updating, while middle age and later life are potentially characterized more by settled dispositions [@alwin1991; @krosnick1989; @eaton2009]. Similarly, highly salient issues, such as views around gay rights in the 2010s; issues that see substantial elite realignments, such as views around the Vietnam conflict in the 1970s [@zaller1992]; and novel issues of public opinion, such as views around vaccines during the Covid-19 pandemic, might be characterized by active updating, while low-salience, established issues are characterized by stability. 

Comparing which of these models better fit a broad range of questions from the General Social Survey’s rotating panels, Kiley and Vaisey [-@kiley2020] found limited evidence of durable change. While the AUM was preferred for the majority of items, the total amount of durable change detected on these items was small. A substantial minority of questions (39 percent) favored the SDM, meaning they showed no evidence of durable change. Questions with more evidence of durable change included highly salient issues like gay marriage and “public” issues like partisan identification and religious service attendance. There was also more evidence of durable change among early adults (people ages 18-30) than the rest of the population. Overall, the researchers concluded that “results ultimately suggest that real, persistent attitude change is an uncommon phenomenon among adults'' [@kiley2020: p. 500; see also @vaisey2021]. This lack of durable change is consistent with other findings that cohort replacement plays a larger role than period effects in explaining differences in personal culture [@vaisey2016].

However, the claim that change is a relatively infrequent phenomenon has been difficult to square with research identifying durable individual change across a number of cultural dimensions, such as morality [@brocic2021], trust [@mewes2021], and concerns about immigration [@kratz2021], as a result of social experiences. Similarly, there are longitudinal studies showing that cues from political elites can change a person’s position on specific policy issues [@slothuus2021; @zaller1992] and that changes in a person’s close contacts and acquaintances can change their attitudes on group-related politics [@gelman2021; @dellaposta2018]. Given how often we observe people change their personal culture, it is hard to accept that adults do not change.

Drawing on these findings, Lersch challenged the SDM and AUM as a “needless dichotomy” [-@lersch2023] and proposed a “Life Course Adaption Model” (LCAM), which draws on the life course perspective and models personal culture over the course of a panel as a linear trajectory that varies for each respondent. Lersch’s work addresses two shortcomings of the AUM and SDM comparison. First, the AUM as outlined by Kiley and Vaisey assumes that change can only occur as a Markovian process, in which responses to personal culture items at time $t$ are a function only of responses at time $t-1$. But early life experiences might indirectly shape peoples’ personal culture as they move into new social roles or environments, even if they do not affect personal culture right away. For example, experiences in childhood might not affect views about family structure until much later in life when people create their own families. The LCAM allows responses at earlier points than $t-1$ to influence responses at time $t$. 

Second, Kiley and Vaisey only adjudicated the AUM and SDM on three-wave panel data covering four years, which might not be a long enough time span to adjudicate the two models they outline. In contrast, Lersch tests the LCAM against the AUM and SDM on panels covering a much longer time frame (between 3 and 36 years) with more waves (between 3 and 18 waves), allowing for a greater possibility of detecting durable change.

In his tournament of models, Lersch finds broad support for the LCAM in data from five countries, with the LCAM favored over the AUM and SDM on 297 of 428 questions he analyzed and the SDM favored over the other two models on 112 items, with the remaining questions producing inconclusive results (no questions preferred the AUM). He concludes that “new experiences over the life course, which expose individuals to new social environments, including novel interactions, and different sets of resources, can persistently move individuals’ personal culture in novel directions” [@lersch2023: p. 243-244].

Despite differences in how these researchers interpret their findings, the empirical results are not far apart. Kiley and Vaisey [-@kiley2020] found that the majority of questions they explored favored the AUM, meaning they showed some evidence of durable change among adults, a finding broadly compatible with Lersch’s results. At the same time, while Lersch [-@lersch2023] found that the LCAM was preferred in the majority of cases he examined, durable change does not account for a large proportion of variance explained on these items. On average, people changed only about .07 standard deviations over 10 years. Again, this is consistent with Kiley and Vaisey’s assertion that durable change is often small. And a quarter of the items Lersch studied still preferred the SDM, suggesting that even with looser assumptions about the process of change, more cases, and more waves, many questions show no evidence of durable change. Relatedly, in most studies purporting to find evidence of durable change in peoples’ personal culture, the change is often quite small. For example, Brocic and Miles [-@brocic2021] find that completing graduate degrees in humanities, arts, and social sciences shift peoples’ moral relativism only about .2 standard deviations on average compared to people with no college degree, and this is the largest effect they identify. And studies of aggregate change show that, even on items where cohort effects dominate, there is always evidence that some people change over time [@vaisey2016]. In other words, despite different interpretations, the results of previous work are in many ways highly consonant.

## Quantification, Not Adjudication

The shortcoming of current debates is that they focus on whether people *ever* change, a question that is not theoretically interesting since the answer is almost always “yes.”  As noted previously, if researchers observe enough people for long enough, they will likely observe some of them making durable changes. A lack of evidence for change in one question might simply indicate that the question was not asked at a clear enough level of resolution, for a long enough time frame, or to enough people to find evidence of change. Conversely, finding evidence of change tells us little about how much change is present in a population and how meaningful that change is for social differences.

A more theoretically productive approach is to start with a model that assumes that, in the window of a panel survey, some people might remain stable, some people might change a little, and some people might undergo substantial changes in personal culture. Lersch’s LCAM formally does this in modeling each person as following their own trajectory over time. But rather than adjudicating whether this model fits the data better than one in which people never change –-asking the question of whether change ever exists-– researchers should quantify the amount that people change over time and compare that change to the differences that exist between people at the beginning of the panel. In other words, rather than ask, “do people change?” we should ask, “what are the relative contributions of change and initial differences toward explaining differences in personal culture?”

In this approach, there are two numbers that each reflect a combination of relevant theoretical processes. First, interpersonal differences at the start of the survey, what Lersch calls “early imprinting,” reflects the accumulated experiences of people prior to entering the panel survey that shape how they answer questions. While this is typically associated with a formative period or a settled disposition (hence the name early imprinting), it could also reflect transformative experiences that happen at any age, so long as they occurred before the start of the panel and leave a lasting effect on how people answer questions. For example, among people already retired upon entering the panel, their “early imprinting” might reflect this life course transition. To some extent, then, this number also reflects variation across people in relevant social roles or statuses at the start of the panel that are important in shaping dispositions.

The second set of processes, what Lersch calls “persistent change” or “adaption” and Kiley and Vaisey call “active updating,” reflects durable changes in personal culture over time and is captured by the amount of intrapersonal linear change over the course of the survey. Lersch suggests that these changes are brought on by social triggers such as moving into a new environment or adopting new social roles. They could also reflect the diffusion of new culture across social networks, changes in social network composition (or knowledge about who is in a person's network), new cues from political elites or other cultural leaders, the emergence of new issues in politics or culture, or other large-scale social shifts.

A third number, often called “residual variance” or “fluctuation,” necessarily accounts for the remainder of variance in peoples’ responses over time. This non-durable change potentially emerges for a variety of reasons. For example, people might not have clear personal cultural dispositions as measured by surveys. Instead, they might internalize a broad set of considerations and construct an opinion or attitude in the context of the survey interview, with different considerations coming to the forefront of their cognition each interview [@tourangeau2000; @zaller1992; @feldman1992]. This variance in considerations can produce variance in responses even if peoples’ “dispositions” --the full set of considerations they could draw on-- remains stable over time. This variance could also include true errors of measurement, such as respondents misunderstanding the question, respondents or administrators selecting the wrong response, or responses getting coded incorrectly. While this third process reflects important aspects of personal culture, it does not directly touch on the ongoing debate outlined here. Because of this we quantify this component but focus principally on the other two.

We assert that, at the theoretical level, the existence of these two principal processes --interpersonal differences and intrapersonal change-- is indisputable, and that they are, in many ways, related. If intrapersonal change occurs at any point in the life course, it will inevitably produce interpersonal differences by the time people enter a panel survey. Similarly, unless people are fully socialized early in the life course and never deviate from these dispositions (a proposition against which there is substantial evidence), then we expect to observe intrapersonal change in some proportion of the population whenever we survey them over time. 

Furthermore, these numbers are not necessarily useful in isolation or without a sense of time. What is important for the theoretical debate is the relative contribution of these two components in explaining cultural differences in a population. In other words, when we look at a population of adults over some time frame, is there enough intrapersonal change relative to interpersonal difference to view it as an important explanation for why people demonstrate cultural differences? That is a question that can only be answered by quantifying the two processes.

However, the real value in quantification lies not in adjudicating past debates, but in opening up avenues of exploration that were limited under previous approaches. For example, classifying questions as either displaying durable change or not obscures differences among questions where there is evidence of durable change, but the relative importance of durable change is different. In quantifying the relative contributions of interpersonal differences and intrapersonal change, researchers can compare the relative importance of these processes in explaining differences across societies, across groups within the same population, across items in the same survey, and across time.

## Expectations

By shifting the question and approach, we expect to see broad similarities across data sets previously used in this analysis.  First, we expect to observe some level of intrapersonal change in personal culture for most measures of personal culture in each data set. However, previous studies found a substantial number of questions (likely between 25 and 35 percent) where there was no evidence of durable change. Consistent with this, we expect to observe significant variation in the proportion of variance explained by intrapersonal change, with some questions showing very little variance explained by intrapersonal change.

Second, we anticipate that the amount of variance explained by intrapersonal change will generally be small relative to differences that exist between people when they start the survey. This assertion should not be controversial, regardless of which theoretical process we expect to be operative for a particular question. Panel surveys do not cover the full span of peoples’ lives, so even if intrapersonal change plays a large role in explaining cultural differences, all the intrapersonal change observed in a panel survey covering even 20 years is likely not enough to account for the full variance of interpersonal differences.

To this point, our expectations are driven principally by what was observed in previous studies. However, theoretical perspectives differ in the amount of variance they expect to be attributable to intrapersonal change, and it is here that quantification can provide a new perspective on existing questions. “Formative period” and cohort theories that expect early-life socialization to be a strong force in adults’ personal culture would require that the amount of variance attributable to intrapersonal change would be insufficient to explain interpersonal differences even when it is extrapolated to a longer time frame. In contrast, other perspectives that attribute more explanatory power to the roles that people currently occupy would expect intrapersonal change to explain a rather large component of variance when extrapolated. While the assumption that the amount of change we observe during the panel could be extrapolated beyond the panel window is not necessarily valid, we believe it is the most valid starting point.

Third, we expect the amount of variance to be explained by each process –interpersonal differences and intrapersonal change– to depend on features of the question and panel. Most obviously, we anticipate that the longer we observe respondents, the greater the degree of resolution at which we observe them (i.e., the number of response options for a particular question), and the larger the sample is, the greater share of variance will be explained by intrapersonal change. Because the Life Course Adaption model (and the various theoretical perspectives that underlie it) suggests that social experiences drive changes of personal culture, and because social experiences are a function of time, observing people for a longer period increases the chance that each of them will undergo some experience that changes their personal culture.

# Analytical Strategy

## Data

We use data from 7 nationally representative panel surveys from Australia, Germany, Great Britain, Switzerland, and the United States (summarized in Table 1), combining all the data files used in previous work [@kiley2020; @lersch2023].[^1] These studies cover a long period of time (the range of surveys spans from 1968 to 2021), with 609 personal culture items that capture attitudes, beliefs, values, self-assessments, self-descriptions, and behaviors [@alwin2007]. We restricted the sample such that individuals between the ages of 18 and 79 are included without further elimination, and in all surveys, we used all possible cases for which respondents provided responses. In the end, the analyses that follow rely on a cross-cultural sample with a cross-domain set of items to capture the breadth of individual personal culture. Appendix A documents the list of all variables used in the upcoming analyses.

[^1]: For more information on these data sources, see [@data_gss; @data_hilda; @data_bhps; @data_ukhls; @data_soep; @data_shp; @data_psid].

```{r, message = FALSE}

library(tidyverse)
library(knitr)
library(kableExtra)

table <- tibble(
  Country = c("Australia", "Germany", "Great Britian", "Great Britian", "Switzerland", "United States", "United States"),
  Survey = c("Household, Income and Labor Dynamics in Australia (HILDA)",
             "Socio-Economic Panel Study (SOEP)",
             "British Household Panel Survey (BHPS)",
             "Understanding Society/UK Household Longitudinal Study (UKHLS)",
             "Swiss Household Panel (SHP)",
             "General Social Survey (GSS)",
             "Panel Study of Income Dynamics (PSID)"),
  Period = c("2001-2021", "1984-2020", "1991-2008", "2009-2020", "1999-2019", "2006-2012", "1968-2019"),
  Outcomes = c(30, 122, 83, 66, 77, 183, 48)
)

kableExtra::kable(table, 
                  format = "latex", 
                  caption = "The Description of the Data Sources") |> 
  kable_styling(latex_options = c("scale_down", "HOLD_position"))

```

## Life Course Adaption Model

We start with Lersch’s [-@lersch2023] Life Course Adaption Model, which formalizes survey responses at time $t$ as a function of individual-level random intercepts and slopes for survey age, what is often called a mixed-effects growth curve model. This model formally assumes a set of propositions about change that reflect the theoretical debate to this point. First, consistent with the settled disposition model, it assumes that people start the survey with cultural differences, modeled as random intercepts for each respondent. Second, it assumes that people change over time, taking the form of random slopes for each respondent as a linear function of time. Third, it assumes that people deviate around this baseline randomly over time, reflecting “fluctuation” or short-term non-persistent change. Finally, it assumes that people draw on earlier time points in reacting to new events. Formally, this can be written as

$$
y_{i, t} = \beta_0 + u_{0, i} + \beta_1 x_t + u_{1, i} x_{i, t} + \epsilon_{i, t}
$$

where $\beta_0$ is the average intercept, $u_{0, i}$ is the random intercept for individuals, $\beta_1 x_t$ is the average slope, $u_{1, i} x_{i, t}$ is the random slope for individuals, and $\epsilon_{i, t}$ is the random error term that captures transitory disturbances.

As a first step in our analysis, we fit LCAMs to each of the 609 measures of personal culture outlined above. We then use these models to derive two measures of systematic variance that correspond to the two collections of theoretical processes outlined above: interpersonal differences at baseline and intrapersonal change over time. First, to measure the importance of interpersonal differences, we calculate the sum of the squared residuals derived from measuring each respondent at the midpoint of their age trajectory.[^midpoint] We then divide this value by the total sum of squares and subtract it from from 1, producing a value we denote as $\tau$, or the proportion of variance explained by interpersonal differences. This number is equivalent to an $R^2$ value generated by using random intercepts for all respondents. Second, we calculate the sum of squared residuals generated by additionally accounting for the age trajectory. We divide this number by the total sum of squares and subtract it from 1. We then subtract $\tau$ from this value. This gives us a value we denote as $\delta$, the additional proportion of variance explained by intrapersonal change. 

These two numbers reflect the total systematic or "explained" variance of the model that can be attributed to either differences across people or linear change over the course of the panel.[^explained] Our principal outcome of interest is the proportion of systematic variance accounted for by intrapersonal change, which we calculate as

$$
\frac{\delta}{\tau + \delta}
$$

[^midpoint]: We use the scale midpoint, rather than the actual difference at baseline, to better capture the differences between people independent of their age trajectories.

[^explained]: This does not mean that other factors might not explain additional variance if they were included in the model. For example, peoples' responses about religious beliefs and participation might systematically vary by the day of the week or month in which they are surveyed. This variation is moved into the "random" or non-systematic component of the model for us. 

We can think of this number as akin to an Intraclass Correlation Coefficient focused only on the systematic components of the model while ignoring residual variance that might be attributable to measurement error or non-persisting change [see also the variance decomposition of period and cohort by @vaisey2016]. The higher the number is, the more variance in responses is attributable to intrapersonal change during the panel. The lower the number, the less intrapersonal change seems to play a large role in explaining cultural differences during the panel. 

Because this quotient summarizes variance and therefore has no natural referent, it is hard to say at which values we observe "a lot" of intrapersonal change. But summarizing questions in this way allows us to compare the relative importance of our two broad processes across a range of questions that might have very different levels of non-systematic variance, giving us a baseline sense of the relative prevalence of these two processes for each question (something not achievable under the tournament of models approach), and allowing us to compare which kinds of questions show more or less intrapersonal change relative to interpersonal differences. While the tournament of models approach allowed researchers to say whether questions in general showed evidence of updating, this approach allows us to make more specific claims about the prevalence of these theoretical processes for particular questions relative to each other.

As a second step in our analysis, we use ordinary least squares regression to model the proportion of variance explained by intrapersonal change as a function of features of the question and panel. This allows us to explore how these features affect the prevalence of intrapersonal change. These covariates include the number of response options (measured as 2, 3-5, and more than 5), the panel, the log of the number of participants, the number of waves observed, the time since the question was first asked, and the total span of time the question was asked.

# Results

Our results proceed in three parts. We first explore the overall pattern of systematic variance explained by interpersonal differences at baseline and intrapersonal change over time across the seven panels. Next, we analyze the association between the relative proportion of systematic variance explained by intrapersonal change as a function of question, panel, and sample features. Finally, we highlight a few specific comparisons to illustrate the value of quantification relative to adjudication for further exploring theoretical questions.

## Variance Decomposition

Figure 1 plots the proportion of systematic variance attributable to interpersonal differences (“between”) and intrapersonal change (“within”) for each of the 609 questions analyzed, plotted separately by panel. To broadly summarize the results, all panels show a similar range of the proportion of systematic variance attributable to intrapersonal change. Across panels, the median proportion of systematic variance attributable to intrapersonal change is 0.094 (mean of 0.098), with an interquartile range of 0.053 to 0.139. All panels include questions that show essentially no systematic variance attributable to intrapersonal change and all panels include questions where more than 20 percent of systematic variance is attributable to intrapersonal change. 

Across all questions, interpersonal differences explain a much larger share of the systematic variance in responses than intrapersonal change. Again, this is to be expected. Interpersonal differences capture not just pre-adult socialization, but all accumulated experiences up to the start of the panel that might have a bearing on personal culture. 

To the extent that there are differences across the panels, the PSID has the highest mean (.120) and median (.135) proportion of systematic variance attributable to intrapersonal change. While we cannot disentangle features of the sample from features of the questions asked to each sample, the specific samples for many PSID questions have lower average ages than those from other panels. To the extent that younger respondents might be more likely to make durable changes of opinion, this higher explanatory power of intrapersonal change might reflect the distinct age profile of respondents in this sample. At the other end, the GSS has the lowest mean (.073) and median (.069) proportion of variance attributable to intrapersonal change. This potentially reflects the fact that the GSS observes people for a shorter duration, on average, than the other panels. If, consistent with life course adaption theories, people are more likely to make significant cultural changes the longer we observe them, then duration likely affects the amount of variance attributable to intrapersonal change (a point we explore more below). However, the GSS results are still highly consistent with results from the other panels. 

While there are some differences between panels, these differences are small compared to the differences within panels. For about 6 percent of items explored here, more than 20 percent of systematic variance is attributable to intrapersonal change. These questions tend to ask about external referents (e.g., confidence in specific government leaders or political parties), life satisfaction, or current financial position. At the other end, questions about religious identification, views on gender roles, and support for civil liberties tend to have very low proportion of variance attributable to intrapersonal change.

In contrast to the tournament of models approach, quantifying change this way allows us to explore variation in the relative importance of intrapersonal change across questions that all both show evidence of change. For example, Kiley and Vaisey [-@kiley2020] found that confidence in the press and confidence in religious leaders were both characterized by active updating, our results show that intrapersonal change is much more important for explaining variance in confidence in the press (0.164) than confidence in religion (0.049). 

Appendix A presents the estimated proportion of variance attributable to interpersonal differences and intrapersonal change, as well as the proportion of residual variance, for each question analyzed. In general, interpersonal differences are almost always the largest component of the total variance and tend to explain between 55 and 70 percent of total variance, while intrapersonal change is always the smallest, typically explaining between 3 and 8 percent of total variance. Residual variance tends to account for between 22 and 37 percent of variance (though on several questions residual variance is greater than 50 percent).

As noted above, the substantive importance of intrapersonal change for understanding cultural differences depends on a range of factors, including how long the panel runs, whether assumptions about linear change hold, and whether the panel is capturing a distinctly turbulent period or a distinctly stable period for the item under question. However, if we assume that the period under observation is “typical” for a question –-not a time of extremely heightened (or lowered) sensitivity or change-– then it does not seem realistic for this process of intrapersonal change to explain a large share of interpersonal differences for many of the questions we observe. 

## Meta-Analysis

Figure 2 plots the results from an ordinary least squares regression of proportion of systematic variance attributable to intrapersonal change as a function of question, panel, and sample features. These models also include fixed effects for panels (not shown), so coefficients reflect the association between a particular sample or question characteristic within a panel.[^paneleffects]

[^paneleffects]: KK NOTE: Can we summarize these panel fixed effects at all? 

Figure 2 shows that the more response options respondents are given and the larger the sample, the larger the proportion of systematic variance is attributable to intrapersonal change. We interpret these coefficients as suggesting greater resolution on a question makes it easier to detect and model change. The earlier a question was asked, even net of how long it was been asked, the lower the proportion of systematic variance is attributable to intrapersonal change.

The most notable result in the meta analysis is the negative coefficient attached to the span of years covered by the question. The longer a question is observed, the less systematic variance is attributable to intrapersonal change. Theories of personal cultural change that link changes in personal culture to social experiences, including the LCAM, suggest that the longer we observe respondents, the more likely people are to undergo potentially transformative experiences and therefore the more variance would be explained by intrapersonal change. The fact that we find a negative coefficient attached to this variable is a challenge to that assumption.

As a further test of this coefficient, we compared the proportion of systematic variance attributable to intrapersonal change when using the full duration of a panel compared to when when we dropped the final wave and therefore reduced the total duration of observation for the question. If the coefficient reflects a true negative effect of duration on systematic variance attributable to intrapersonal change, we should see that same effect within questions. Results from this analysis are presented in Appendix B and contradict the coefficient from the regression model. The longer we observe the same question, the higher proportion of systematic variance is attributable to intrapersonal change. We interpret this combination of findings as suggesting that the kinds of questions asked for a longer time period tend to demonstrate less intrapersonal change than questions asked over shorter periods, rather than a true function of time.

This reinforces the explanation for why the GSS displays the lowest average proportion of systematic variance attributable to intrapersonal change: its shorter duration. [KK NOTE: Does this inform any other panel differences? Does the PSID, BHPS or UKHLS have longer durations on average?]

## A Case Study: College Education

While these broad results help speak to previous debates, they mostly re-frame previous findings without extending the theoretical debate. To better illustrate the value of our approach for addressing different theoretical questions, we turn to a specific empirical example: the relative importance of intrapersonal change and interpersonal difference for explaining cultural differences by education levels. 

Previous work has established a positive relationship between education and attitude stability, especially on issues related to American politics. This stability is often attributed to education facilitating "chronic knowledge" -- a general understanding of the domain of American politics, including the positions held by major parties and political figures and how issues relate to one another at a logical or socio-logical level [@zaller1992; @alvarez2002; @boutyline2017]. Because they have more knowledge of American politics, people with a college education are better able to consistently connect the considerations in their cognition with the answer choices they are presented with in a survey.

This work has tended to focus on the fact that college-educated Americans give responses that are less likely to be affected by measurement error or short-term influences [@zaller1992; @alwin2007]. However, this focus on the difference in the non-systematic (residual) component of variance across groups obscures the fact that the relative importance of different components of systematic variance might differ across these two groups as well.

It could be the case that education, chronic knowledge, and attention to the political sphere makes intrapersonal change more important in explaining differences among people with higher levels of education. For example, these people might be more likely to make systematic changes in response to the emergence of new information, new issues, or political realignments, while those without chronic knowledge display more variance around their baselines without their baselines changing much [@zaller1992]. In contrast, it could be the case that the stability demonstrated by educated respondents reflects the fact that they have already formed durable opinions and are relatively closed off to new information. College could serve as a formative experience that solidifies personal culture for people who attend. But among people who do not attend college, later life experiences might prove more important in forming or changing personal culture, as these experiences potentially provide information that college-educated peers received earlier. 

To test these competing propositions, we use the models presented above to calculate the total systematic variance for people with at least a bachelor's degree and people with less than a bachelor's degree at wave 1[^wave] of the General Social Survey's panels.[^missing] We focus on the GSS because it contains the largest number of questions tapping general political dispositions, which is the domain where education has proven particularly relevant for understanding attitude stability. The GSS also covers a particularly turbulent window of American politics from 2006 to 2014. This window covers the start of the Great Recession, debates about federal intervention in and regulation of Wall Street, the election of Barack Obama as the first black president in U.S. history, debates about the role of the federal government in the health care sector, the emergence of the Tea Party, and political realignment and clarification on the issue of gay marriage, among other topics.

[^wave]: A small number of respondents report different highest degrees at wave 1, wave 2, and wave 3. Some of this is due to measurement error, and some of it is due to a small number of people obtaining a higher degree during the four years of the panel. Estimating the panel with highest at wave 1 or highest degree reported across the panel produces functionally identical results.

[^missing]: We estimate the models presented in [earlier figure] using the full sample, but only use respondents who have information on their highest degree to generate the results presented in figures [these figures]. In total, ### respondents (%%% percent) reported no degree information.

Figure XXX plots the distribution of differences between people with less than a bachelor's degree and people with at least a bachelor's degree at wave 1 of the panel. Values greater than 0 would indicate that intrapersonal change is more important in explaining the responses of college graduates than those of people without a college degree, while values less than 0 would indicate that intrapersonal change is less important for explaining variation among people with a college degree than those without a college degree.

There is a clear pattern in [figure above]: for more than 80 percent of GSS items explored here, intrapersonal change is a larger component of systematic variance for people without a college degree than for people with a college degree (they have negative values in the figure). While most of these differences are small in absolute terms (less than 2 percentage points), several are greater than 5 percentage points. Given the distribution observed in Figure [the earlier one] showing that the systematic variance attributable to intrapersonal change averages [average across all questions], a 5 percentage point difference between groups is quite substantial. 

To more clearly illustrate some of these differences, we highlight eight questions designed to tap general political dispositions: partisan identification (Democrat vs. Republican) and ideological identification (liberal vs. conservative) on seven-point scales; four questions about the government's role in improving the condition of the poor, paying people's medical bills, giving special treatment to blacks, and doing things that private businesses could do, measured on five-point scales; a question about whether the government should do more to reduce income differences, measured on a seven-point scale; and one question about whether blacks should be given preferences in hiring, measured on a five-point scale. Estimates for the proportion of systematic variance attributable to intrapersonal change on these eight questions, for both education groups, are presented in Figure XXX.

On all eight questions presented in Figure XXX, intrapersonal change explains less systematic variance for people with a college degree than for people without a college degree. There are also large differences across questions on the amount of systematic variance attributable to intrapersonal change for both groups. For example, on the question of whether the government should try to solve more problems or leave those problems to be solved by private businesses ("government do more or less"), less than one percent of the systematic variance is attributable to intrapersonal change for both groups. In other words, while people might vacillate on this question at random (37 percent of variance is residual for this question), there is functionally no evidence that people make systematic changes of opinion on this issue during the GSS panel.

In contrast, partisan identification and political ideology both show a larger proportion of systematic variance attributable to intrapersonal change, with quite large differences between people with a college degree and those without. Compared to the other questions explored here, change as adults (and, potentially, their experiences as adults) plays a much larger role in explaining why people report the partisan identification and political ideology they do. And this is particularly true for respondents who do not have a college degree. The amount of systematic variance in ideological identification that is attributable to intrapersonal change among non-college educated respondents is almost four times that of college-educated respondents. 

We note here that this meaningful difference in the proportion of systematic variance attributable to intrapersonal change across education groups and across questions would not be detectable using previous methods. For partisan identification and political ideology, both college-educated respondents and people with less than college degree would likely favor the AUM or LCAM over the SDM because they show evidence of some members of the population making intrapersonal change. In other words, using the tournament of models approach obscures the fact that the relative importance of intrapersonal change in explaining differences on these two items is very different across these two groups, that intrapersonal change explains more systematic variance for ideological identification than partisan identification for people without a college degree, or that there appears to be more durable change on questions of affirmative action than on questions of government aid to black Americans. 

It is important to note here that these differences should not be interpreted as causal effects of attending college on the importance of social experiences for cultural change. These results are potentially confounded by age, social class, race, gender, and other factors that explain selection into higher education. We simply use these estimates as a starting point for a broader discussion.

# Discussion

This paper set out to clarify previous findings about the prevalence of durable change in personal culture -- the attitudes, beliefs, values, preferences, and behaviors that manifest at the individual level. Relying on an approach that quantified the amount of systematic variance attributable to either intrapersonal change or interpersonal differences at baseline, we found a consistent pattern across data sets previously used in this debate. We see evidence of intrapersonal change over time on almost all questions, with some measures of personal culture displaying quite high levels of intrapersonal change relative to the amount of variance between people. On some questions, such as the factors that affect one’s life satisfaction or attitudes about government benefits, it is highly plausible that differences in peoples’ experiences as adults are the primary explanation for differences in personal culture.

However, the amount of intrapersonal change is often small relative to interpersonal differences (less than 10 percent on average across all surveys). On many questions, the amount of systematic variance attributable to intrapersonal change is essentially zero. These include questions about civil liberties, abortion, generalized trust, or sense of civic duty. On these questions and the large number of others with a low proportion of systematic variance attributable to intrapersonal change, there does not appear to be sufficient cultural change during adulthood for experiences and transitions between social locations to be the primary explanation for why adults end up with different personal cultures. For these items, the explanation for personal cultural differences seems to reside principally in early-life experiences.

More important than adjudicating past debates, however, we show that the approach developed in this paper can adjudicate novel theoretical questions about the conditions in which change or stability is more prevalent and more important for explaining cultural differences within a population. As an example, we showed that the role of intrapersonal change in explaining cultural differences varies significantly by question and by education. Intrapersonal change plays almost no role in explaining general sentiments about the government's role in the economy, while it plays a much larger role in explaining ideological and partisan identification. At the same time, across issues, intrapersonal change is much more important for explaining cultural variation among people without a college degree than for people with a college degree. While this pattern might be confounded by age and other factors that affect college completion, it potentially suggests that college-going locks in personal culture in a way that other experiences do not, making adult experiences less influential for political dispositions for college graduates compared to the rest of the population.

## Implications for Cultural Sociology

To this point, the theoretical debate in cultural sociology has focused on the existence of personal cultural change among adults and its relative importance. We believe the "existence" debate is theoretically misguided. Personal cultural change will always exist to some degree in an adult population, and no viable theoretical model expects just one process --stability or change-- to characterize a population. Our results reinforce that assertion, finding a mixture of interpersonal differences and intrapersonal change for all items. Even when we observe very low levels of systematic variance attributable to intrapersonal change for particular questions, this is potentially attributable to insufficient data, insufficient resolution on the underlying belief, or insufficient duration of observation.

A more reasonable theoretical debate pushed by previous work is the relative importance of these two processes for explaining cultural variation. While change might always be prevalent, its impact on social differences might be small or large, and our approach has been designed to clarify the contribution of intrapersonal change to such differences. It is difficult to draw a single conclusion on this question across a wide range of questions measured on different scales and over different durations of time, but the overall picture, with important caveats, suggests that intrapersonal change during adulthood is not the dominant explanation for cultural differences among adults for most items we measure in surveys. There simply is not enough evidence of intrapersonal change on these items relative to the amount of difference we observe across people.

To be clear, none of our results should be interpreted as suggesting that adults do not change or are incapable of changing. On all questions, some members of the sample change their personal culture at least a little bit. Our results also should not be taken to imply that intrapersonal change is not important or consequential. Small numbers of changers can have important consequences in the aggregate. A shift from 49 percent supporting gay marriage to 51 percent supporting it can have significant political ramifications, even as most respondents remain fixed in their preferences. And our results should not detract from work trying to estimate the mechanisms through which personal cultural differences --however small-- emerge among adults. Just because change is small does not mean the mechanisms that produce it are not important to understand.

Our results should be read principally as suggesting that if researchers want to explain why people differ in their personal culture, it is imperative to look at the conditions in which they were raised and the experiences that happened before they reached adulthood. To the extent that we want to explain what makes adults different, focusing attention on peoples’ transition between social roles, their selection into and out of organizational environments, and changes in their social networks, as much sociological work does, appears to explain a much smaller proportion of their differences than the formative experiences that happen early in life.

In this way, our findings echo a range of recent causal inference work that suggests selection effects, rather than treatment effects, seem to explain the majority of differences in personal culture between people in different social roles or positions [@campbell2016; @wodtke2018]. While some people clearly change over time as they move into new social roles or into new social environments, this amount of change does not appear to be sufficient to explain what are often substantial differences between people in different roles. Our results suggest that, to the extent we observe differences in personal culture between people in different social roles, such as parenthood, having children, education, or working in positions of authority [@longest2013], or between people in different occupations [@weeden2005], our prior expectation should be that this difference is selection, rather than treatment.

Our results do not speak to exactly why intrapersonal change lacks significant explanatory power in understanding culture. First, it could be the case that adults are exposed to opportunities for change at the same rate, but become less willing to change. Or it could be that opportunities for durable change or the social conditions that facilitate durable change are just extremely uncommon. Changes in social structures, the life course, or cultural expectations could make intrapersonal change a much more consequential explanation for cultural differences.

Furthermore, the results we present related to education and political views suggest that the relative importance of different theoretical processes might differ by group. In line with life course perspectives, the results presented here suggest that importance of different experiences for cultural change are potentially contingent on previous experiences. For example, to the extent that we want to explain why people have different personal culture, the answer is likely different for people who have a college degree and those who do not. Mid-life experiences appear to have more significant implications for the personal culture for people without a college degree than for people with a college degree. 

## Implications for Social Change

Because intrapersonal change in personal culture does not appear to be a driving force in cultural differences among adults, our results suggest that processes of social and cultural change at the aggregate level are likely driven more by cohort replacement, one group of people with certain cultural dispositions exiting the population and a different set of people with a potentially different set of cultural dispositions entering the population, than by contemporaneous social conditions [@vaisey2016; @underwood2022]. This is likely true for both large-scale change in societies over time as well as changes within organizations, political parties, occupations, and other social aggregates.

However, because our results suggest that the role of formative experiences compared to contemporaneous environments differs by a person's level of education, changes in educational attainment over time, as well as changes in the social structuring of the life course might be shifting the relative importance of these two factors over time. If we observe further differences across socio-demographic groups like education, social class, and race/ethnicity in the relative importance of these processes, this would argue for a greater incorporation of the interaction of demographic processes with processes of individual cultural change into our explanations of aggregate cultural change.

## Implications for Survey Measurement

A major takeaway from the meta-analysis is that greater response resolution –-the number of response options survey respondents are given-– is associated with a greater share of variance being explained by within-individual change. Again, it is possible that this reflects different kinds of issues being measured with different kinds of scales. But it could also suggest that within-person change, when it happens, is relatively small –-moving from “agree” to “strongly agree,” rather than “strongly agree” to “strongly disagree,” for example-– and therefore easier to detect with greater resolution. 

Our results continue to highlight the value of long-running panel surveys for adjudicating different theories of culture that cannot be evaluated with cross-sectional data alone. Across our results, residual fluctuations in responses account for more than four times as much variance, on average, as linear persistent change, meaning that most change in responses is not systematic. Under these circumstances, extending panels beyond two waves provides significantly more leverage to understand how and why personal culture changes. 

Additionally, we echo previous calls for more studies –-including panel studies-– of the socialization processes of young people. Our results suggest that most of the variance between adults emerges before people turn 18 or relatively soon after. In limiting our empirical perspective to experiences that happen after that point, social scientists potentially focus on questions that, while important, lack the capacity to explain major cultural differences.

## Limitations

The approach outlined here generates estimates of how much variance in a population is due to each of two broad processes: intrapersonal change and interpersonal differences at baseline. It does not quantify the proportion of people who “change.” When we observe intrapersonal change in a question, we cannot be sure whether this is driven by many people making small changes or a few people making large changes. Adjudicating this question will require different models that distinguish different kinds of change.

Our results and interpretations also hinge on how we have defined change and the assumptions built into the Life Course Adaption Model. This model treats cultural trajectories as varying linear slopes for each respondent, assuming that change is a linear function of time. This assumption likely does not hold true in practice. Change over time likely takes non-linear and discontinuous forms. People might jump from one “stable” disposition to another, for example, or experience a discontinuous "turning point" that upsets a stable trajectory. This assumption is in part a limitation of the data, as most questions are only observed for three to six waves with a large number of missing cases. Panels with more waves might allow for more flexible models of change that do not require this assumption of linearity.

Similarly, in assuming a linear trajectory for change, our approach assumes that durable change is unidirectional. To the extent that people make durable change as a result of a social transition and then return to a previous position later in life as a result of a second transition (marriage and divorce, for example), this is captured as residual within-person variance, rather than one of our two broad theoretical processes. This is a potentially reasonable assumption on short panels --change that does not last two years is hard to classify as durable-- but it might not be true in longer panels. Other researchers with more flexible definitions of "durable change" might potentially identify more durable change even in the same population.

Finally, while we examine an incredible broad array of measures of personal culture, from assessments about the relative importance of different features when buying a new car and policy preferences to religious beliefs and core values, we are limited by the kinds of questions that are asked in panel surveys and the contexts in which these surveys were administered (Australia, Germany, Switzerland, the United Kingdom and the United States). This means our findings tend to reflect questions about general (national) politics, gender roles, immigration and race relations, and general well being. Our data does not cover some dimensions of culture that might be relevant for other aspects of sociological exploration (artistic tastes, leisure activities, and time use, for example).

## Implications for Future Research

We believe the approach outlined here can push past the “needless dichotomy” implicit in the question of whether people change or not and address more theoretically informed questions about the conditions under which the relative importance of intrapersonal change and interpersonal differences matter for explaining cultural differences. Characterizing questions as displaying change or not can only take researchers so far, but the question of why some questions demonstrate more change while others do not, or why some groups are characterized by more stability than others, has the potential to weigh in on a broader range of theoretical debates. We hope researchers move toward those questions.

While we focused on education as a determinant of the explanatory power of intrapersonal change, our approach can readily be extended to other socio-demographic categories that might affect intrapersonal change --race/ethnicity, religiosity, social class, gender, and political affiliation, for example-- or their intersections. Researchers with different data might be able to evaluate the relative importance of intrapersonal change and interpersonal differences across contexts as well, either national contexts if the same questions are administered across countries, or organizational contexts if data traces students within schools, workers within workplaces, or congregation members within congregations.

The wide variation across questions in the proportion of systematic variation attributable to intrapersonal change also calls for a more systematic exploration for why some forms of culture are characterized by stability while others are characterized by change, and what the consequences of these differences for the organization of beliefs in the public might be.

# References
